---
title: "digiRhythm demo"
output: html_document
---

This notebook showcases the use of the 
[digiRhythm package](https://cran.r-project.org/web/packages/digiRhythm/index.html).

is an R package for the analysis of animal activity during night and day. The package was written by Dr. Roland Nasser and an description of the project can be found at [DigiRhythm](https://www.agroscope.admin.ch/agroscope/en/home/topics/economics-technology/smart-farming/digirhythm.html)

In this package we will explore a sample dataset of this package and showcase the different plots that the pacakge offers.

## Setup

Load the package

```{r, eval=F, message=F, warning=F}
install.packages("devtools")
install.packages('digiRhythm', dependencies = T)
```

## Load the data

Here we are loading an example dataset provided by the package. The package provides promises to fetch the
data: so for this notebook we will use this promise to avoid downloading the dataset. 

The dataset contains the observations of one animal and tracks it's activity index and steps over time: the observations are recorded evers 15 minutes. 

```{r}
library(digiRhythm)

# Select an input file containing position recordings over time.
filename <- here::here('data', '516b_2.csv')

#The columns that we are interested in
colstoread <- c("Date", "Time", "Motion Index", 'Steps') 

# Reading the activity data from the csv file
df <- import_raw_activity_data(
  filename = filename,
  skipLines = 7,
  act.cols.names = colstoread,
  sampling = 15)
```

## Inspect the data
The dataset can be inspected by the `is_dgm_friendly` function: it checks whether the observations have the expected attributes and frequencies.

```{r}
is_dgm_friendly(df, verbose=T)
```

Metadata such as the time period that was monitored is provided by the function `df_act_info`.

```{r}
df_act_info(df)
```

## Visualization

Specific visualizations are also provided, with an option to directly save the figure to a file.

### Activity over time

First, the `actogram` function shows activity along the day for all dates at the same time, as a heatmap. This function allows to specify a column to visualize, a date range to include. When we are satisfied with the figure, we can save it by giving a file path with the `save` option. 

For this animal we can see that the activity is very regular and peeks in the morning and evening.

```{r}
# Define time range to include in figures
start <- '2020-04-30'
end <- '2020-06-14'

p <- actogram(df,
         activity = 'Motion.Index',
         activity_alias = 'Motion Index',
         start = start,
         end = end,
         save = NULL)

```

`daily_average_activity` provides an alternative visualization where observation are grouped by the time of day
the curve shows the mean activity on a given time of the day. The function allows for chosing a time period for 
the grouping of the observations: it looks at the average activity during a day in a given time interval.

Again we can see that in the chosen time interval the activity peeks in the morning and in the evening.

```{r}
p <- daily_average_activity(df,
         activity = 'Motion.Index',
         activity_alias = 'Motion Index',
         start = start,
         end = end,
         save = NULL)

```

All plotting functions in digiRhythm return ggplot2 objects. 
We can therefore customize the way we display them: here below we are adding a title and print out the start and end date of the observation that were grouped and aggregated.

```{r}
library(ggplot2)
title <- sprintf(
  "Daily average over the period %s to %s",
  start,
  end
  )
p +
  theme_minimal() +
  ggtitle(title)
```

### Periodicity

Initially the time between observations is 15 Minutes: we can confirm this by printing out the periodictiy of the dataset.

```{r}
dgm_periodicity(df)
```

### Diurnality

The `diurnality` function computes and plots the diurnality index over the year, computed as $\frac{A_d - A_n}{A_d+A_n}$ where $A_d$ and $A_n$ is the total activity during day and night, respectively. A positive diurnality index means that the animal is more active during the day than night.

```{r}
diur <- diurnality(df, activity = 'Steps')
```
